---
title: "Final Proejct"
format: html
editor: visual
---

# Load packages

```{r}
#| message: false
#| message: false

library(tidymodels)
library(tidyverse)
library(recipes)  # step_rm, step_impute_* etc.
library(lubridate)
library(ranger)
library(rsample)
```

# Loading and inspecting the training and testing files

```{r}
#| message: false
#| message: false

# list.files("../data/training", full.names = TRUE)
# list.files("../data/testing", full.names = TRUE)

```

## **Read all 6 CSVs**

```{r}
#| message: false
#| message: false

train_meta  = read_csv("../data/training/training_meta.csv")
train_soil  = read_csv("../data/training/training_soil.csv")
train_trait = read_csv("../data/training/training_trait.csv")

test_meta   = read_csv("../data/testing/testing_meta.csv")
test_soil   = read_csv("../data/testing/testing_soil.csv")
test_sub    = read_csv("../data/testing/testing_submission.csv")
```

# Inspect column names

```{r}
#| message: false
#| message: false

# names(train_meta)
# names(train_soil)
# names(train_trait)
# 
# names(test_meta)
# names(test_soil)
# names(test_sub)
```

## **Join meta and soil into trait/submission**

```{r}
#| message: false
#| message: false

# Combine META and SOIL for training and testing data
train_env = train_meta %>%
  left_join(train_soil, by = c("year", "site"))

test_env = test_meta %>%
  left_join(test_soil, by = c("year", "site"))

# Add env info to trait
train_df = train_trait %>%
  left_join(train_env, by = c("year", "site"))

test_df = test_sub %>%
  left_join(test_env, by = c("year", "site"))

# checks 
glimpse(train_df)
glimpse(test_df)

# EDA
summary(train_df$yield_mg_ha)
table(train_df$year)
table(train_df$site)
```

## **Site-year grouping variable for CV use**

```{r}
#| message: false
#| message: false

train_df = train_df %>%
  mutate(site_year = paste(site, year, sep = "_"))
train_df

test_df = test_df %>%
  mutate(site_year = paste(site, year, sep = "_"))
test_df
```

## **Tidymodels and set-up of recipe**

```{r}
#| message: false
#| message: false

# convert character predictors to factors
train_df = train_df %>%
  mutate(
    site          = factor(site),
    hybrid        = factor(hybrid),
    previous_crop = factor(previous_crop),
    site_year     = factor(site_year)
  )

test_df = test_df %>%
  mutate(
    site          = factor(site),
    hybrid        = factor(hybrid),
    previous_crop = factor(previous_crop),
    site_year     = factor(site_year)
  )

# date parsing helper funvtion (month.day.year to Date)
parse_mdy_year = function(x, year) {
  # x have "5/5/14" format, so I coerce using lubridate::mdy with paste to date 
  suppressWarnings(
    lubridate::mdy(paste(x, year))
  )
}

# count of predictor columns BEFORE recipe steps
n_pred = train_df %>%
  select(-yield_mg_ha, -site_year) %>%
  ncol()

n_pred

```

# Define a recipe using yield_mg_ha as outcome

```{r}
#| message: false
#| message: false

corn_rec = recipe(
  yield_mg_ha ~ year + site + replicate + block + hybrid +
    date_planted + date_harvested +
    grain_moisture +
    previous_crop + longitude + latitude +
    soilpH + om_pct + soilk_ppm + soilp_ppm,
  data = train_df
) %>%
  # Convert character predictors to factors
  step_string2factor(all_nominal_predictors()) %>%
  
  # Date variables from the character date columns
  step_mutate(
    planted_date   = mdy(date_planted),
    harvested_date = mdy(date_harvested)
  ) %>%
  
  # Engineer numeric date features
  step_mutate(
    doy_planted      = yday(planted_date),
    doy_harvested    = yday(harvested_date),
    harvest_interval = as.numeric(harvested_date - planted_date)
  ) %>%
  
  # Remove original string and intermediate Date columns
  step_rm(date_planted, date_harvested, planted_date, harvested_date) %>%
  
  # Impute missing numeric and categorical data
  step_impute_median(all_numeric_predictors()) %>%   # numeric median imputation[web:37][web:43]
  step_impute_mode(all_nominal_predictors()) %>%     # mode imputation for factors[web:40]
  
  # Dummy categorical predictors encoded
  step_dummy(all_nominal_predictors())

corn_rec

```

## **Create grouped resampling**

```{r}
#| message: false
#| message: false

set.seed(123)

cv_folds = group_vfold_cv(
  train_df,
  group = site_year,
  v = 5
)

corn_rec
cv_folds

```

## **Tuned XGBoost model with grouped CV**

```{r}
#| message: false
#| message: false

# XGBoost spec with tunable hyperparameters
xgb_spec = boost_tree(
  trees          = 1000,
  tree_depth     = tune(),
  learn_rate     = tune(),
  loss_reduction = tune(),
  min_n          = tune(),
  sample_size    = tune(),
  mtry           = tune()
) %>%
  set_mode("regression") %>%
  set_engine("xgboost")

xgb_spec
```

# XGBoost workflow

```{r}
#| message: false
#| message: false

xgb_wf = workflow() %>%
  add_model(xgb_spec) %>%
  add_recipe(corn_rec)

xgb_wf
```

## **Tuning grid for** XGBoost

```{r}
#| message: false
#| message: false

set.seed(123)

# Use a Latin space‑filling over a sensible range.
xgb_grid = grid_space_filling(
  tree_depth(),
  learn_rate(range = c(-3, -0.5)),
  loss_reduction(),
  min_n(),
  sample_size = sample_prop(),
  mtry(range = c(5, min(50, n_pred))),  # use the n_pred counted from train_df
  size = 30
)

xgb_grid
```

# Tune XGBoost with grouped CV

```{r}
set.seed(123)

xgb_tuned = tune_grid(
  xgb_wf,
  resamples = cv_folds,
  grid      = xgb_grid,
  metrics   = metric_set(rmse, rsq),
  control   = control_grid(save_pred = TRUE)
)

xgb_tuned

collect_metrics(xgb_tuned) %>%
  filter(.metric == "rmse") %>%
  arrange(mean)

```

# XGB best parameters

```{r}
best_xgb = select_best(xgb_tuned, metric = "rmse")
best_xgb

# Finalize the workflow
final_xgb_wf = finalize_workflow(xgb_wf, best_xgb)
```

# Fit on all training data and predict 2024

```{r}
final_xgb_fit = fit(final_xgb_wf, data = train_df)

xgb_preds = predict(final_xgb_fit, new_data = test_df) %>%
  bind_cols(test_df %>% select(year, site, hybrid)) %>%
  rename(pred_yield_mg_ha = .pred)

xgb_preds
```

## 
