---
title: "Final Project"
format: html
editor: visual
---

# Load packages

```{r}
#| message: false
#| message: false

library(tidymodels)
library(tidyverse)
library(recipes)  # step_rm to work
library(lubridate)
library(ranger)
library(rsample)
```

# Loading and inspecting the training and testing data files

```{r}
#| message: false
#| message: false

# list.files("../data/training", full.names = TRUE)
# list.files("../data/testing", full.names = TRUE)

```

## **Read all CSV files**
```{r}
#| message: false
#| warning: false

train_meta  = read_csv("../data/training/training_meta.csv")
train_soil  = read_csv("../data/training/training_soil.csv")
train_trait = read_csv("../data/training/training_trait.csv")

test_meta   = read_csv("../data/testing/testing_meta.csv")
test_soil   = read_csv("../data/testing/testing_soil.csv")
test_sub    = read_csv("../data/testing/testing_submission.csv")
```

## Clean site names and join all files

```{r}
#| message: false
#| warning: false

clean_site = function(s) {
  s <- str_remove(s, "_\\d{4}$")         
  s <- str_remove(s, "[abc]$")            
  s <- str_remove(s, "-\\s*(Dry|Early|Late)")   
  s <- str_remove(s, "\\s*-?\\s*rep\\s*\\d+")  
  s <- str_remove(s, "_[12]$")            
  str_trim(s)
}

# Clean and deduplicate training meta
train_meta_clean = train_meta %>%
  mutate(site = clean_site(site)) %>%
  group_by(year, site) %>%
  summarise(
    previous_crop = first(previous_crop),
    longitude     = mean(longitude, na.rm = TRUE),
    latitude      = mean(latitude,  na.rm = TRUE),
    .groups       = "drop"
  )

# Clean and deduplicate training soil
train_soil_clean = train_soil %>%
  mutate(site = clean_site(site)) %>%
  group_by(year, site) %>%
  summarise(
    soilpH    = mean(soilpH,    na.rm = TRUE),
    om_pct    = mean(om_pct,    na.rm = TRUE),
    soilk_ppm = mean(soilk_ppm, na.rm = TRUE),
    soilp_ppm = mean(soilp_ppm, na.rm = TRUE),
    .groups   = "drop"
  )

# Clean testing soil
test_soil_clean = test_soil %>%
  mutate(site = clean_site(site)) %>%
  group_by(year, site) %>%
  summarise(
    soilpH    = mean(soilpH,    na.rm = TRUE),
    om_pct    = mean(om_pct,    na.rm = TRUE),
    soilk_ppm = mean(soilk_ppm, na.rm = TRUE),
    soilp_ppm = mean(soilp_ppm, na.rm = TRUE),
    .groups   = "drop"
  )

# Clean site in trait before joining
train_trait_clean = train_trait %>%
  mutate(site = clean_site(site))

# Build training dataset: trait + meta + soil
train_df = train_trait_clean %>%
  left_join(train_meta_clean,  by = c("year", "site")) %>%
  left_join(train_soil_clean,  by = c("year", "site"))

# build testing dataset: submission + meta + soil
test_df = test_sub %>%
  left_join(test_meta,         by = c("year", "site")) %>%
  left_join(test_soil_clean,   by = c("year", "site"))

#  Check results 
cat("Training rows:", nrow(train_df), "\n")
cat("Testing rows:",  nrow(test_df),  "\n")

cat("\nTraining NAs per column:\n")
train_df %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "column", values_to = "n_NA") %>%
  filter(n_NA > 0) %>%
  print()

cat("\nTesting NAs per column:\n")
test_df %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "column", values_to = "n_NA") %>%
  filter(n_NA > 0) %>%
  print()
```

## Add site_year grouping variable

```{r}
#| message: false
#| warning: false

train_df = train_df %>%
  mutate(site_year = paste(site, year, sep = "_"))

test_df = test_df %>%
  mutate(site_year = paste(site, year, sep = "_"))
```

## Convert character columns to factors

```{r}
#| message: false
#| warning: false

train_df = train_df %>%
  mutate(
    site          = factor(site),
    hybrid        = factor(hybrid),
    previous_crop = factor(previous_crop),
    site_year     = factor(site_year)
  )

test_df = test_df %>%
  mutate(
    site          = factor(site),
    hybrid        = factor(hybrid),
    previous_crop = factor(previous_crop),
    site_year     = factor(site_year)
  )
```

## Count predictors (needed for mtry in XGBoost grid)

```{r}
n_pred = train_df %>%
  select(-yield_mg_ha, -site_year) %>%
  ncol()

cat("Number of predictors:", n_pred, "\n")
```
## Pull weather data using nasapower

```{r}
#| message: false
#| warning: false
#| eval: false   

library(nasapower)

# Get unique site-year-lat-lon from BOTH training and testing meta
# We need weather for every site-year we have coordinates for
sites_train = train_meta_clean %>%
  filter(!is.na(latitude), !is.na(longitude)) %>%
  select(year, site, latitude, longitude) %>%
  distinct()

sites_test = test_meta %>%
  select(year, site, latitude, longitude) %>%
  filter(!is.na(latitude), !is.na(longitude)) %>%
  distinct()

all_sites = bind_rows(sites_train, sites_test) %>%
  distinct(year, site, .keep_all = TRUE)

cat("Total site-years to pull weather for:", nrow(all_sites), "\n")

# Function to pull growing season weather for one site-year
# Corn growing season = May (day 121) to October (day 304)
pull_weather = function(year, lat, lon, site) {
  
  tryCatch({
    raw = get_power(
      community  = "AG",
      lonlat     = c(lon, lat),
      pars       = c("T2M_MAX",   # daily max temperature (C)
                     "T2M_MIN",   # daily min temperature (C)  
                     "PRECTOTCORR", # precipitation (mm/day)
                     "ALLSKY_SFC_SW_DWN"), # solar radiation (MJ/m2/day)
      dates      = c(paste0(year, "-05-01"),   # May 1
                     paste0(year, "-10-31")),  # Oct 31
      temporal_api = "daily"
    )
    
    raw %>%
      mutate(
        site = site,
        year = year,
        # Growing degree days for corn (base 10C, max 30C)
        tmax_capped = pmin(T2M_MAX, 30),
        tmin_capped = pmax(T2M_MIN, 10),
        gdd = pmax((tmax_capped + tmin_capped) / 2 - 10, 0)
      ) %>%
      group_by(year, site) %>%
      summarise(
        # Temperature summaries
        tmean_season   = mean((T2M_MAX + T2M_MIN) / 2, na.rm = TRUE),
        tmax_mean      = mean(T2M_MAX, na.rm = TRUE),
        tmin_mean      = mean(T2M_MIN, na.rm = TRUE),
        # Heat stress: days where max temp > 35C
        heat_stress_days = sum(T2M_MAX > 35, na.rm = TRUE),
        # Growing degree days (accumulated)
        gdd_total      = sum(gdd, na.rm = TRUE),
        # Precipitation
        precip_total   = sum(PRECTOTCORR, na.rm = TRUE),
        # Solar radiation
        srad_mean      = mean(ALLSKY_SFC_SW_DWN, na.rm = TRUE),
        .groups = "drop"
      )
    
  }, error = function(e) {
    message("Failed for site=", site, " year=", year, ": ", e$message)
    return(NULL)
  })
}

# Pull weather for all site-years
cat("Pulling weather data... this takes ~5-10 minutes\n")

weather_list <- vector("list", nrow(all_sites))

for (i in seq_len(nrow(all_sites))) {
  cat("Pulling", i, "of", nrow(all_sites), 
      "| site:", all_sites$site[i], 
      "| year:", all_sites$year[i], "\n")
  
  weather_list[[i]] <- pull_weather(
    year = all_sites$year[i],
    lat  = all_sites$latitude[i],
    lon  = all_sites$longitude[i],
    site = all_sites$site[i]
  )
  
 
  Sys.sleep(0.5)
}

# Combine all results
weather_df <- bind_rows(weather_list)

cat("Weather data pulled successfully!\n")
cat("Rows:", nrow(weather_df), "\n")
cat("Columns:", ncol(weather_df), "\n")

# Save to file
write_csv(weather_df, "../data/weather_season_summary.csv")
cat("Saved to ../data/weather_season_summary.csv\n")
```

## Load weather and join to training and testing data

```{r}
#| message: false
#| warning: false

# Load the saved weather file
weather_df = read_csv("../data/weather_season_summary.csv")

# Join weather to training data
train_df = train_df %>%
  left_join(weather_df, by = c("year", "site"))

# Join weather to testing data
test_df = test_df %>%
  left_join(weather_df, by = c("year", "site"))

cat("Training columns after weather join:", ncol(train_df), "\n")
cat("Testing columns after weather join:", ncol(test_df), "\n")

# Check weather NAs
cat("\nWeather NAs in training:\n")
train_df %>%
  select(tmean_season, tmax_mean, tmin_mean, 
         heat_stress_days, gdd_total, precip_total, srad_mean) %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "column", values_to = "n_NA") %>%
  print()

# Update n_pred after adding weather columns
n_pred = train_df %>%
  select(-yield_mg_ha, -site_year) %>%
  ncol()

cat("\nUpdated number of predictors:", n_pred, "\n")
```


# Define a recipe using yield_mg_ha as the outcome

```{r}
#| message: false
#| message: false

corn_rec = recipe(
  yield_mg_ha ~ year + site + replicate + block + hybrid +
    date_planted + date_harvested +
    grain_moisture +
    previous_crop + longitude + latitude +
    soilpH + om_pct + soilk_ppm + soilp_ppm +
    tmean_season + tmax_mean + tmin_mean +
    heat_stress_days + gdd_total + precip_total + srad_mean,
  data = train_df
) %>%
  # Convert character predictors to factors
  step_string2factor(all_nominal_predictors()) %>%
  
  # Date variables from the character date columns
  step_mutate(
    planted_date   = mdy(date_planted),
    harvested_date = mdy(date_harvested)
  ) %>%
  
  # Engineer numeric date features
  step_mutate(
    doy_planted      = yday(planted_date),
    doy_harvested    = yday(harvested_date),
    harvest_interval = as.numeric(harvested_date - planted_date)
  ) %>%
  
  # Remove original string and intermediate Date columns
  step_rm(date_planted, date_harvested, planted_date, harvested_date) %>%
  
  # Impute missing numeric and categorical data
  step_impute_median(all_numeric_predictors()) %>%   # numeric median imputation[web:37][web:43]
  step_impute_mode(all_nominal_predictors()) %>%     # mode imputation for factors[web:40]
  
  # Dummy categorical predictors encoded
  step_dummy(all_nominal_predictors())

corn_rec

```

## **Create grouped resampling**

```{r}
#| message: false
#| message: false

set.seed(123)

cv_folds = group_vfold_cv(
  train_df,
  group = site_year,
  v = 5
)

corn_rec
cv_folds

```

## **Tuned XGBoost model with grouped CV**

```{r}
#| message: false
#| message: false

# XGBoost spec with tunable hyperparameters
xgb_spec = boost_tree(
  trees          = 1000,
  tree_depth     = tune(),
  learn_rate     = tune(),
  loss_reduction = tune(),
  min_n          = tune(),
  sample_size    = tune(),
  mtry           = tune()
) %>%
  set_mode("regression") %>%
  set_engine("xgboost")

xgb_spec
```

# XGBoost workflow

```{r}
#| message: false
#| message: false

xgb_wf = workflow() %>%
  add_model(xgb_spec) %>%
  add_recipe(corn_rec)

xgb_wf
```

## **Tuning grid for** XGBoost

```{r}
#| message: false
#| message: false

set.seed(123)

# Use Latin space‑filling over a range in the tuning
xgb_grid = grid_space_filling(
  tree_depth(),
  learn_rate(range = c(-3, -0.5)),
  loss_reduction(),
  min_n(),
  sample_size = sample_prop(),
  mtry(range = c(5, min(50, n_pred))),  # use the n_pred counted from train_df
  size = 30
)

xgb_grid
```

# Tune XGBoost with grouped CV

```{r}
set.seed(123)

xgb_tuned = tune_grid(
  xgb_wf,
  resamples = cv_folds,
  grid      = xgb_grid,
  metrics   = metric_set(rmse, rsq),
  control   = control_grid(save_pred = TRUE)
)

xgb_tuned

collect_metrics(xgb_tuned) %>%
  filter(.metric == "rmse") %>%
  arrange(mean)

```

# XGB best parameters

```{r}
best_xgb = select_best(xgb_tuned, metric = "rmse")
best_xgb

# Finalize the workflow
final_xgb_wf = finalize_workflow(xgb_wf, best_xgb)
```

# Predicted vs observed on CV predictions
```{r}
xgb_tuned %>%
  collect_predictions() %>%
  inner_join(select_best(xgb_tuned, metric = "rmse")) %>%
  ggplot(aes(x = yield_mg_ha, y = .pred)) +
  geom_point(alpha = 0.3) +
  geom_abline(color = "red") +
  geom_smooth(method = "lm") +
  labs(title = "XGBoost: Predicted vs Observed",
       x = "Observed Yield (mg/ha)",
       y = "Predicted Yield (mg/ha)")
```



# Fit on all training data and predict 2024

```{r}
final_xgb_fit = fit(final_xgb_wf, data = train_df)

xgb_preds = predict(final_xgb_fit, new_data = test_df) %>%
  bind_cols(test_df %>% select(year, site, hybrid)) %>%
  rename(pred_yield_mg_ha = .pred)

xgb_preds
```

## Vip
```{r}
library(vip)
final_xgb_fit %>%
  extract_fit_parsnip() %>%
  vip(num_features = 15) +
  labs(title = "XGBoost Variable Importance")
```

```{r}
# Save final predictions to submission file
test_sub %>%
  mutate(yield_mg_ha = xgb_preds$pred_yield_mg_ha) %>%
  write_csv("../output/testing_submission_filled_xgb.csv")
```




























# machine learning models: GLMNet


```{r}
#| message: false
#| warning: false

library(glmnet)
library(tidyverse)
library(tidymodels)
# GLMNet has only 2 hyperparameters:
# penalty  = how much to shrink coefficients (lambda)
# mixture  = 0 is pure Ridge, 1 is pure Lasso, 0.5 is Elastic Net
glmnet_spec = linear_reg(
  penalty = tune(),
  mixture = tune()
) %>%
  set_engine("glmnet") %>%
  set_mode("regression")

glmnet_spec
```

## GLMNet recipe

```{r}
#| message: false
#| warning: false

# GLMNet REQUIRES normalized predictors (unlike XGBoost)
# So we add step_normalize() here
glmnet_rec = recipe(
  yield_mg_ha ~ year + site + replicate + block + hybrid +
    date_planted + date_harvested +
    grain_moisture +
    previous_crop + longitude + latitude +
    soilpH + om_pct + soilk_ppm + soilp_ppm +
    tmean_season + tmax_mean + tmin_mean +
    heat_stress_days + gdd_total + precip_total + srad_mean,
  data = train_df
) %>%
  # Convert character predictors to factors
  step_string2factor(all_nominal_predictors()) %>%
  
  # Parse date columns
  step_mutate(
    planted_date   = mdy(date_planted),
    harvested_date = mdy(date_harvested)
  ) %>%
  
  # Engineer date features
  step_mutate(
    doy_planted      = yday(planted_date),
    doy_harvested    = yday(harvested_date),
    harvest_interval = as.numeric(harvested_date - planted_date)
  ) %>%
  
  # Remove original date columns
  step_rm(date_planted, date_harvested, planted_date, harvested_date) %>%
  
  # Impute missing values
  step_impute_median(all_numeric_predictors()) %>%
  step_impute_mode(all_nominal_predictors()) %>%
  
  # Dummy encode categorical variables
  step_dummy(all_nominal_predictors()) %>%
  
  # IMPORTANT: GLMNet needs normalized predictors
  step_normalize(all_numeric_predictors())

glmnet_rec
```

## GLMNet workflow

```{r}
#| message: false
#| warning: false

glmnet_wf = workflow() %>%
  add_model(glmnet_spec) %>%
  add_recipe(glmnet_rec)

glmnet_wf
```

## GLMNet tuning grid

```{r}
#| message: false
#| warning: false

set.seed(488)

# Simple grid - only 2 hyperparameters so we can try many combinations
glmnet_grid = grid_regular(
  penalty(range = c(-4, 0)),   # from 0.0001 to 1
  mixture(range = c(0, 1)),    # from pure Ridge to pure Lasso
  levels = 10                  # 10x10 = 100 combinations
)

cat("GLMNet grid size:", nrow(glmnet_grid), "combinations\n")
glmnet_grid
```

## Tune GLMNet with grouped CV

```{r}
set.seed(488)

cv_folds = group_vfold_cv(
  train_df,
  group = site_year,
  v = 5
)
```

```{r}
#| message: false
#| warning: false

# Use same cv_folds as XGBoost for fair comparison
set.seed(488)

# Parallel processing to speed up
library(doParallel)
registerDoParallel(cores = parallel::detectCores() - 1)

glmnet_tuned = tune_grid(
  glmnet_wf,
  resamples = cv_folds,
  grid      = glmnet_grid,
  metrics   = metric_set(rmse, rsq),
  control   = control_grid(save_pred = TRUE)
)

stopImplicitCluster()

cat("GLMNet tuning complete!\n")

# Show best results
collect_metrics(glmnet_tuned) %>%
  filter(.metric == "rmse") %>%
  arrange(mean) %>%
  head(10)
```

## Select best GLMNet model

```{r}
#| message: false
#| warning: false

# Select best model based on lowest RMSE
best_glmnet = glmnet_tuned %>%
  select_best(metric = "rmse")

best_glmnet

# Finalize workflow with best parameters
final_glmnet_wf = finalize_workflow(glmnet_wf, best_glmnet)
final_glmnet_wf
```

## GLMNet predicted vs observed plot

```{r}
#| message: false
#| warning: false

# Collect CV predictions for best model
glmnet_tuned %>%
  collect_predictions() %>%
  inner_join(best_glmnet) %>%
  ggplot(aes(x = yield_mg_ha, y = .pred)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_abline(color = "red", linewidth = 1) +
  geom_smooth(method = "lm", color = "blue") +
  labs(
    title = "GLMNet: Predicted vs Observed Yield",
    x     = "Observed Yield (mg/ha)",
    y     = "Predicted Yield (mg/ha)"
  ) +
  theme_bw()
```

## GLMNet performance metrics

```{r}
#| message: false
#| warning: false

# Show RMSE and R2 for best GLMNet model
glmnet_tuned %>%
  collect_metrics() %>%
  filter(.metric %in% c("rmse", "rsq")) %>%
  inner_join(best_glmnet) %>%
  select(.metric, mean, std_err) %>%
  print()
```

## Fit GLMNet on all training data and predict 2024

```{r}
#| message: false
#| warning: false

# Fit final model on full training data
final_glmnet_fit = fit(final_glmnet_wf, data = train_df)

# Predict 2024 test data
glmnet_preds = predict(final_glmnet_fit, new_data = test_df) %>%
  bind_cols(test_df %>% select(year, site, hybrid)) %>%
  rename(pred_yield_mg_ha = .pred)

glmnet_preds
```

## Save GLMNet predictions to submission file

```{r}
#| message: false
#| warning: false

# Save GLMNet predictions
test_sub %>%
  mutate(yield_mg_ha = glmnet_preds$pred_yield_mg_ha) %>%
  write_csv("../output/testing_submission_glmnet.csv")

cat("GLMNet predictions saved!\n")
```

#Shiny app

```{r}
# HOW TO RUN:
# 1. Save this file as app.R in your code/ folder
# 2. Open in RStudio and click "Run App"

library(shiny)
library(tidyverse)
library(ggplot2)

# ── Set working directory to project root ─────────────────────
setwd("C:/Users/ag78178/OneDrive - University of Georgia/Desktop/Amit_Learning/Amit_PhD_USA-data/Coursework UGA PhD/Spring 2026/CRSS8030/2026dsa_finalproject_group10")

# ── Load all data ─────────────────────────────────────────────
train_trait <- read_csv("data/training/training_trait.csv",
                        show_col_types = FALSE)

train_soil  <- read_csv("data/training/training_soil.csv",
                        show_col_types = FALSE) %>%
  mutate(site = str_remove(site, "_\\d{4}$")) %>%
  group_by(year, site) %>%
  summarise(soilpH    = mean(soilpH,    na.rm = TRUE),
            om_pct    = mean(om_pct,    na.rm = TRUE),
            soilk_ppm = mean(soilk_ppm, na.rm = TRUE),
            soilp_ppm = mean(soilp_ppm, na.rm = TRUE),
            .groups   = "drop")

weather <- read_csv("data/weather_season_summary.csv",
                    show_col_types = FALSE)

val_preds   <- read_csv("Sapelo_output/output/xgb_internal_validation_predictions.csv",
                        show_col_types = FALSE)

val_metrics <- read_csv("Sapelo_output/output/xgb_internal_validation_metrics.csv",
                        show_col_types = FALSE)

# ── Pre-compute summaries ─────────────────────────────────────
yield_year <- train_trait %>%
  filter(!is.na(yield_mg_ha)) %>%
  group_by(year) %>%
  summarise(mean_yield = mean(yield_mg_ha),
            sd_yield   = sd(yield_mg_ha),
            .groups    = "drop")

soil_long <- train_soil %>%
  pivot_longer(cols      = c(soilpH, om_pct, soilk_ppm, soilp_ppm),
               names_to  = "variable",
               values_to = "value") %>%
  filter(!is.na(value))

weather_long <- weather %>%
  pivot_longer(cols      = c(tmean_season, precip_total,
                              gdd_total, heat_stress_days),
               names_to  = "variable",
               values_to = "value")

rmse_val <- val_metrics %>%
  filter(.metric == "rmse") %>%
  pull(.estimate) %>% round(3)

r2_val <- val_metrics %>%
  filter(.metric == "rsq") %>%
  pull(.estimate) %>% round(3)

# ── UI ────────────────────────────────────────────────────────
ui <- fluidPage(

  tags$head(tags$style(HTML("
    body { background-color:#f4f6f9; font-family:Georgia,serif; }
    h2   { color:#2c5f2e; border-bottom:2px solid #2c5f2e;
           padding-bottom:8px; margin-bottom:20px; }
    .metric-box {
      background:#2c5f2e; color:white; border-radius:10px;
      padding:20px; text-align:center; margin:8px;
    }
    .metric-box h3 { font-size:2.2em; margin:0; color:#f8c200; }
    .metric-box p  { margin:0; font-size:1em; }
    .well { background:#fff; border:1px solid #dde; border-radius:8px; }
    .nav-tabs>li.active>a {
      background-color:#2c5f2e !important; color:white !important;
    }
    .nav-tabs>li>a { color:#2c5f2e; font-weight:bold; }
  "))),

  # Header
  div(style="background:#2c5f2e;color:white;padding:20px 30px;margin-bottom:20px;",
    h1("🌽 Corn Yield Prediction Dashboard",
       style="margin:0;font-size:1.8em;"),
    p("Team 10 | Amit Godara & Taiwo Owolanke | 2026 DSA Final Project",
      style="margin:5px 0 0 0;opacity:0.85;")
  ),

  # Metric boxes
  fluidRow(
    column(3, div(class="metric-box",
      h3("164K"), p("Training Observations"))),
    column(3, div(class="metric-box",
      h3("22"), p("Predictors Used"))),
    column(3, div(class="metric-box",
      h3(r2_val), p("XGBoost Validation R²"))),
    column(3, div(class="metric-box",
      h3(rmse_val), p("RMSE (Mg/ha)")))
  ),

  br(),

  tabsetPanel(

    # ── Tab 1: Yield EDA ────────────────────────────────────
    tabPanel("📊 Yield EDA",
      br(),
      fluidRow(
        column(3,
          wellPanel(
            h4("Controls", style="color:#2c5f2e;"),
            selectInput("eda_type", "Plot type:",
              choices = c(
                "Mean Yield by Year" = "year",
                "Yield Distribution" = "dist",
                "Yield by Site"      = "site"
              )
            ),
            conditionalPanel("input.eda_type == 'year'",
              checkboxInput("show_sd", "Show error bars", TRUE)
            ),
            conditionalPanel("input.eda_type == 'dist'",
              sliderInput("bins", "Number of bins:", 20, 80, 40)
            ),
            conditionalPanel("input.eda_type == 'site'",
              selectInput("site_sel", "Select site:",
                choices  = sort(unique(train_trait$site)),
                selected = "DEH1"
              )
            )
          )
        ),
        column(9, plotOutput("yield_plot", height = "450px"))
      )
    ),

    # ── Tab 2: Soil EDA ─────────────────────────────────────
    tabPanel("🌱 Soil EDA",
      br(),
      fluidRow(
        column(3,
          wellPanel(
            h4("Controls", style="color:#2c5f2e;"),
            selectInput("soil_var", "Soil variable:",
              choices = c(
                "Soil pH"             = "soilpH",
                "Organic Matter (%)"  = "om_pct",
                "Soil K (ppm)"        = "soilk_ppm",
                "Soil P (ppm)"        = "soilp_ppm"
              )
            ),
            radioButtons("soil_type", "Plot type:",
              choices = c("Histogram"       = "hist",
                          "Boxplot by Year" = "box")
            )
          )
        ),
        column(9, plotOutput("soil_plot", height = "450px"))
      )
    ),

    # ── Tab 3: Weather EDA ──────────────────────────────────
    tabPanel("🌤️ Weather EDA",
      br(),
      fluidRow(
        column(3,
          wellPanel(
            h4("Controls", style="color:#2c5f2e;"),
            selectInput("wx_var", "Weather variable:",
              choices = c(
                "Mean Temperature (°C)"    = "tmean_season",
                "Total Precipitation (mm)" = "precip_total",
                "Growing Degree Days"      = "gdd_total",
                "Heat Stress Days (>35°C)" = "heat_stress_days"
              )
            ),
            sliderInput("wx_years", "Year range:",
              min   = 2014, max   = 2024,
              value = c(2014, 2024), sep = ""
            )
          )
        ),
        column(9, plotOutput("wx_plot", height = "450px"))
      )
    ),

    # ── Tab 4: Variable Importance ──────────────────────────
    tabPanel("🔍 Variable Importance",
      br(),
      h2("XGBoost — Top Predictors of Corn Yield"),
      p("Variables ranked by contribution to the XGBoost model.",
        style = "color:#555;"),
      div(style = "text-align:center;",
        imageOutput("vip_img", height = "500px")
      )
    ),

    # ── Tab 5: Model Performance ────────────────────────────
    tabPanel("📈 Model Performance",
      br(),
      h2("XGBoost — Predicted vs Observed Yield"),
      fluidRow(
        column(3,
          wellPanel(
            h4("Model Metrics", style="color:#2c5f2e;"),
            br(),
            div(style="text-align:center;",
              p("Validation R²"),
              h3(r2_val, style="color:#2c5f2e;margin:0;"),
              hr(),
              p("Validation RMSE"),
              h3(paste0(rmse_val, " Mg/ha"),
                 style="color:#2c5f2e;margin:0;"),
              hr(),
              p("5-fold grouped CV by site-year",
                style="color:#777;font-size:0.85em;")
            )
          )
        ),
        column(9, plotOutput("pred_plot", height = "480px"))
      )
    )

  ) # end tabsetPanel
) # end fluidPage


# ── SERVER ────────────────────────────────────────────────────
server <- function(input, output, session) {

  # Yield plot
  output$yield_plot <- renderPlot({

    if (input$eda_type == "year") {
      p <- ggplot(yield_year,
                  aes(x = factor(year), y = mean_yield)) +
        geom_col(fill = "#2c5f2e", alpha = 0.85, width = 0.7) +
        geom_hline(yintercept = mean(yield_year$mean_yield),
                   linetype = "dashed",
                   color    = "#f8c200",
                   linewidth = 1) +
        labs(title = "Mean Corn Yield by Year (2014-2023)",
             x = "Year", y = "Mean Yield (Mg/ha)") +
        theme_minimal(base_size = 14) +
        theme(plot.title = element_text(color="#2c5f2e", face="bold"))
      if (input$show_sd) {
        p <- p + geom_errorbar(
          aes(ymin = mean_yield - sd_yield,
              ymax = mean_yield + sd_yield),
          width = 0.3, color = "#333")
      }
      p

    } else if (input$eda_type == "dist") {
      train_trait %>%
        filter(!is.na(yield_mg_ha)) %>%
        ggplot(aes(x = yield_mg_ha)) +
        geom_histogram(bins  = input$bins,
                       fill  = "#2c5f2e",
                       color = "white",
                       alpha = 0.85) +
        geom_vline(
          xintercept = mean(train_trait$yield_mg_ha, na.rm=TRUE),
          color      = "#f8c200",
          linewidth  = 1.2,
          linetype   = "dashed") +
        labs(title = "Distribution of Corn Yield — Training Data",
             x = "Yield (Mg/ha)", y = "Count") +
        theme_minimal(base_size = 14) +
        theme(plot.title = element_text(color="#2c5f2e", face="bold"))

    } else {
      train_trait %>%
        filter(site == input$site_sel, !is.na(yield_mg_ha)) %>%
        ggplot(aes(x = factor(year), y = yield_mg_ha)) +
        geom_boxplot(fill  = "#2c5f2e",
                     alpha = 0.7,
                     color = "#1a3d1c") +
        labs(title = paste("Yield by Year at Site:", input$site_sel),
             x = "Year", y = "Yield (Mg/ha)") +
        theme_minimal(base_size = 14) +
        theme(plot.title = element_text(color="#2c5f2e", face="bold"))
    }
  })

  # Soil plot
  output$soil_plot <- renderPlot({
    label <- switch(input$soil_var,
      soilpH    = "Soil pH",
      om_pct    = "Organic Matter (%)",
      soilk_ppm = "Soil K (ppm)",
      soilp_ppm = "Soil P (ppm)"
    )
    dat <- soil_long %>% filter(variable == input$soil_var)

    if (input$soil_type == "hist") {
      ggplot(dat, aes(x = value)) +
        geom_histogram(bins  = 30,
                       fill  = "#5a8f3c",
                       color = "white",
                       alpha = 0.85) +
        labs(title = paste("Distribution of", label),
             x = label, y = "Count") +
        theme_minimal(base_size = 14) +
        theme(plot.title = element_text(color="#2c5f2e", face="bold"))
    } else {
      ggplot(dat, aes(x = factor(year), y = value)) +
        geom_boxplot(fill  = "#5a8f3c",
                     alpha = 0.7,
                     color = "#1a3d1c") +
        labs(title = paste(label, "by Year"),
             x = "Year", y = label) +
        theme_minimal(base_size = 14) +
        theme(plot.title  = element_text(color="#2c5f2e", face="bold"),
              axis.text.x = element_text(angle=45, hjust=1))
    }
  })

  # Weather plot
  output$wx_plot <- renderPlot({
    label <- switch(input$wx_var,
      tmean_season     = "Mean Temperature (°C)",
      precip_total     = "Total Precipitation (mm)",
      gdd_total        = "Growing Degree Days",
      heat_stress_days = "Heat Stress Days (Tmax > 35°C)"
    )
    weather_long %>%
      filter(variable == input$wx_var,
             year     >= input$wx_years[1],
             year     <= input$wx_years[2]) %>%
      group_by(year) %>%
      summarise(mean_val = mean(value, na.rm=TRUE),
                .groups  = "drop") %>%
      ggplot(aes(x = year, y = mean_val)) +
      geom_line(color = "#2c5f2e", linewidth = 1.3) +
      geom_point(color = "#f8c200", size = 4) +
      labs(title = paste("Mean", label, "by Year"),
           x = "Year", y = label) +
      theme_minimal(base_size = 14) +
      theme(plot.title = element_text(color="#2c5f2e", face="bold"))
  })

  # Variable importance image
  output$vip_img <- renderImage({
    list(
      src   = "Sapelo_output/output/png/xgb_variable_importance.png",
      width = "75%",
      style = "display:block;margin:auto;"
    )
  }, deleteFile = FALSE)

  # Predicted vs observed
  output$pred_plot <- renderPlot({
    val_preds %>%
      ggplot(aes(x = yield_mg_ha, y = .pred)) +
      geom_point(alpha = 0.2, color = "#2c5f2e", size = 1) +
      geom_abline(color    = "red",
                  linewidth = 1,
                  linetype  = "dashed") +
      geom_smooth(method    = "lm",
                  color     = "#f8c200",
                  linewidth = 1.2,
                  se        = FALSE) +
      annotate("label",
               x         = 3,
               y         = 14,
               label     = paste0("R² = ", r2_val,
                                  "\nRMSE = ", rmse_val, " Mg/ha"),
               size      = 5,
               color     = "#2c5f2e",
               fontface  = "bold",
               fill      = "white",
               label.size = 0.5) +
      labs(title    = "XGBoost: Predicted vs Observed Corn Yield",
           subtitle = "Grouped 5-fold cross-validation by site-year",
           x = "Observed Yield (Mg/ha)",
           y = "Predicted Yield (Mg/ha)") +
      theme_minimal(base_size = 14) +
      theme(plot.title    = element_text(color="#2c5f2e", face="bold"),
            plot.subtitle = element_text(color="#555"))
  })

} # end server

shinyApp(ui, server)
   
  

```

