---
title: "Final Project Group 10"
format: html
editor: visual
author1: Taiwo Owolanke
author2: Amit Godara
---

```{r} # shared_lib_path <- "/work/crss8030/instructor_data/shared_R_libs"  #class_packages <- c("tidymodels", "tidyverse", "vip", "ranger", "finetune", "parsnip", "reticulate", "xgboost", "doParallel", "lme4", "here")  #install.packages(class_packages, lib = shared_lib_path)}
```

# Locate the installed libraries (important step while running the script in cluster - sapelo2)

```{r} # .libPaths(c("/work/crss8030/instructor_data/shared_R_libs", .libPaths()))}
```

# Load packages/Libraries

```{r}
#| message: false
#| warning: false

library(tidymodels)
library(tidyverse)    # Data wrangling and visualization
library(finetune)     # Additional tuning strategies (e.g racing, ANOVA-based tuning)
library(recipes)  # step_rm to work
library(lubridate)
library(ranger)       # Fast implementation of Random Forests
library(rsample)
library(nasapower)
library(glmnet)
library(vip)          # For plotting variable importance from fitted models
library(xgboost)      # XGBoost implementation in R
library(doParallel)   # For parallel computing (useful during resampling/tuning)
library(dplyr)

library(here)         # Gives the absolute path starting from Rproj, works well with different scripts (R, .rmd or .qmd)
library(quarto)       # helps to export the qmd script to R script
#library(caret)       # Other great library for Machine Learning
library(janitor)
library(stringr)
library(tidyr)

here::here()
```

## **Loading and inspecting the training and testing data files**
```{r}
#| message: false
#| warning: false

# list.files("../data/training", full.names = TRUE)
# list.files("../data/testing", full.names = TRUE)

```

## **Read all CSV files**
```{r}
#| message: false
#| warning: false

train_meta  = read_csv(here("data/training", "training_meta.csv"))
train_soil  = read_csv(here("data/training", "training_soil.csv"))
train_trait = read_csv(here("data/training", "training_trait.csv"))

test_meta   = read_csv(here("data/testing", "testing_meta.csv"))
test_soil   = read_csv(here("data/testing", "testing_soil.csv"))
test_sub    = read_csv(here("data/testing", "testing_submission.csv"))
```

## **Pre-processing**
## Clean site names and join all files
```{r}
#| message: false
#| warning: false

# site extraction functions
extract_site_base <- function(x) {
  x %>%
    str_to_lower() %>%
    str_extract("^[a-z]{2,3}h\\d+") %>%
    str_to_upper()
}

make_site_year <- function(year, site_base) {
  paste(year, site_base, sep = "_")
}

# clean training trait data
train_trait_clean <- train_trait %>%
  clean_names() %>%
  mutate(
    site_base = extract_site_base(site),
    site_year = make_site_year(year, site_base),

    treatment = case_when(
      str_detect(str_to_lower(site), "dry")   ~ "dry",
      str_detect(str_to_lower(site), "early") ~ "early",
      str_detect(str_to_lower(site), "late")  ~ "late",
      TRUE ~ "normal"
    ),

    date_harvested = mdy(date_harvested)
  ) %>%
  filter(!is.na(site_base))   # drop malformed site names early

# clean training meta data at site level only
train_meta_clean <- train_meta %>%
  clean_names() %>%
  mutate(site_base = extract_site_base(site)) %>%
  filter(!is.na(site_base)) %>%
  group_by(site_base) %>%
  summarise(
    previous_crop = first(na.omit(previous_crop)),
    longitude     = mean(longitude, na.rm = TRUE),
    latitude      = mean(latitude, na.rm = TRUE),
    .groups = "drop"
  )

# clean training soil data at site level only
train_soil_clean <- train_soil %>%
  clean_names() %>%
  mutate(site_base = extract_site_base(site)) %>%
  filter(!is.na(site_base)) %>%
  group_by(site_base) %>%
  summarise(
    soilpH    = mean(soilp_h, na.rm = TRUE),
    om_pct    = mean(om_pct, na.rm = TRUE),
    soilk_ppm = mean(soilk_ppm, na.rm = TRUE),
    soilp_ppm = mean(soilp_ppm, na.rm = TRUE),
    .groups = "drop"
  )

# identify valid sites
valid_sites <- train_trait_clean %>%
  distinct(site_base) %>%
  inner_join(train_meta_clean, by = "site_base") %>%
  inner_join(train_soil_clean, by = "site_base")

# join training traits, meta and soil cleaned data
train_df <- train_trait_clean %>%
  semi_join(valid_sites, by = "site_base") %>%  # keep only valid sites
  left_join(train_meta_clean, by = "site_base") %>%
  left_join(train_soil_clean, by = "site_base")

# debug missing sites
missing_sites = train_trait_clean %>%
  distinct(site_base) %>%
  anti_join(train_meta_clean, by = "site_base")

print(missing_sites)

soil_vars <- c("soilpH", "om_pct", "soilk_ppm", "soilp_ppm")

# global medians as named vector
global_medians <- train_df %>%
  summarise(across(all_of(soil_vars), ~ median(.x, na.rm = TRUE))) %>%
  unlist()

train_df <- train_df %>%
  group_by(site_base) %>%
  mutate(across(
    all_of(soil_vars),
    ~ ifelse(
        is.na(.x),
        ifelse(
          all(is.na(.x)),
          global_medians[[cur_column()]],   # fallback to global
          median(.x, na.rm = TRUE)         # site-level median
        ),
        .x
      )
  )) %>%
  ungroup()

train_df <- train_df %>%
  mutate(previous_crop = replace_na(previous_crop, "unknown"))

# drop data having missing harvest dates
train_df <- train_df %>%
  filter(!is.na(date_harvested))

# drop missing sites not in train data and train meta data
train_df <- train_df %>%
  filter(!site_base %in% missing_sites)

na_check <- train_df %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "column", values_to = "n_NA") %>%
  filter(n_NA > 0)

print(na_check)

cat("Training rows:", nrow(train_df), "\n")
glimpse(train_df)
```


```{r}
#| message: false
#| warning: false

# site cleaner and extraction function to match train data
extract_site_base <- function(s) {
  s %>%
    str_to_lower() %>%
    str_extract("^[a-z]{2,3}h\\d+") %>%
    str_to_upper()
}

# clean test soil data
test_soil_clean <- test_soil %>%
  clean_names() %>%
  mutate(site_base = extract_site_base(site)) %>%
  group_by(year, site_base) %>%
  summarise(
    soilpH    = mean(soilp_h, na.rm = TRUE),
    om_pct    = mean(om_pct, na.rm = TRUE),
    soilk_ppm = mean(soilk_ppm, na.rm = TRUE),
    soilp_ppm = mean(soilp_ppm, na.rm = TRUE),
    .groups = "drop"
  )

# clean test meta
test_meta_clean <- test_meta %>%
  clean_names() %>%
  mutate(
    previous_crop = str_to_lower(previous_crop),
    site_base = extract_site_base(site)
  ) %>%
  group_by(year, site_base) %>%
  summarise(
    previous_crop = first(previous_crop),
    longitude = mean(longitude, na.rm = TRUE),
    latitude  = mean(latitude,  na.rm = TRUE),
    .groups = "drop"
  )

# clean test submission data
test_sub_clean <- test_sub %>%
  clean_names() %>%
  mutate(
    site_base = extract_site_base(site)
  )

# build test data set
test_df <- test_sub_clean %>%
  
  # join meta and soil
  left_join(test_meta_clean, by = c("year", "site_base")) %>%
  left_join(test_soil_clean, by = c("year", "site_base")) %>%
  
  # handle missing value
  mutate(
    previous_crop = replace_na(previous_crop, "unknown"),
    
    across(
      c(soilpH, om_pct, soilk_ppm, soilp_ppm),
      ~ replace_na(., mean(., na.rm = TRUE))
    )
  ) %>%
  
  # fill lat/long within site
  group_by(site_base) %>%
  mutate(
    longitude = ifelse(is.na(longitude),
                       mean(longitude, na.rm = TRUE),
                       longitude),
    latitude = ifelse(is.na(latitude),
                      mean(latitude, na.rm = TRUE),
                      latitude)
  ) %>%
  ungroup()

# checks results
cat("Test rows:", nrow(test_df), "\n")

na_check <- test_df %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "column", values_to = "n_NA") %>%
  filter(n_NA > 0)

print(na_check)
```

## Add site_year grouping variable
```{r}
#| message: false
#| warning: false

train_df = train_df %>%
  mutate(site_year = paste(site, year, sep = "_"))

test_df = test_df %>%
  mutate(site_year = paste(site, year, sep = "_"))
```


## Convert character columns to factors
```{r}
#| message: false
#| warning: false

train_df = train_df %>%
  mutate(
    site          = factor(site),
    hybrid        = factor(hybrid),
    previous_crop = factor(previous_crop),
    site_year     = factor(site_year)
  )

test_df = test_df %>%
  mutate(
    site          = factor(site),
    hybrid        = factor(hybrid),
    previous_crop = factor(previous_crop),
    site_year     = factor(site_year)
  )
```


# Distribution of target variable
```{r distribution}
#| message: false
#| warning: false

density_plot <- ggplot() +
  geom_density(data = train_df, 
               aes(x = yield_mg_ha),
               color = "red") +
  # geom_density(data = test_df, 
  #              aes(x = yield_mg_ha),
  #              color = "blue") 
# test data has X variable as NA's, so omitted
  
# lets print the undergoing process in the console, (important when running in the terminal later on)
cat(paste0("Saving...... density plot of train data"))

ggsave(plot = density_plot, 
       path = here("output", "png"),
       filename = "density_plot_train.png",
       height = 6,
       width = 9,
       dpi = 600)
```

## Count predictors (needed for mtry in XGBoost grid)
```{r}
n_pred = train_df %>%
  select(-yield_mg_ha, -site_year) %>%
  ncol()

cat("Number of predictors:", n_pred, "\n")
```

## Pull weather data using nasapower
```{r}
#| message: false
#| warning: false
#| eval: false

clean_site_base <- function(x) {
  x %>%
    str_to_lower() %>%
    str_extract("[a-z]{2,3}h?\\d+[a-z]?") %>%   # handles IAH1, TXH4, NYH1A
    str_to_upper()
}

# Get unique site-year-lat-lon from BOTH training and testing meta
# We need weather for every site-year we have coordinates for
sites_train <- train_df %>%
  transmute(
    year,
    site_base,
    latitude,
    longitude
  ) %>%
  filter(!is.na(latitude), !is.na(longitude)) %>%
  distinct()

sites_test <- test_meta %>%
  clean_names() %>%
  mutate(site_base = clean_site_base(site)) %>%
  transmute(
    year,
    site_base,
    latitude,
    longitude
  ) %>%
  filter(!is.na(latitude), !is.na(longitude)) %>%
  distinct()

all_sites <- bind_rows(sites_train, sites_test) %>%
  distinct(year, site_base, .keep_all = TRUE)

cat("Total site-years:", nrow(all_sites), "\n")

cat("Total site-years to pull weather for:", nrow(all_sites), "\n")

# Function to pull growing season weather for one site-year
# Corn growing season = May (day 121) to October (day 304)
pull_weather = function(year, lat, lon, site) {

  tryCatch({
    raw = get_power(
      community  = "AG",
      lonlat     = c(lon, lat),
      pars       = c("T2M_MAX",   # daily max temperature (C)
                     "T2M_MIN",   # daily min temperature (C)
                     "PRECTOTCORR", # precipitation (mm/day)
                     "ALLSKY_SFC_SW_DWN"), # solar radiation (MJ/m2/day)
      dates      = c(paste0(year, "-05-01"),   # May 1
                     paste0(year, "-10-31")),  # Oct 31
      temporal_api = "daily"
    )

    raw %>%
      mutate(
        site = site,
        year = year,
        # Growing degree days for corn (base 10C, max 30C)
        tmax_capped = pmin(T2M_MAX, 30),
        tmin_capped = pmax(T2M_MIN, 10),
        gdd = pmax((tmax_capped + tmin_capped) / 2 - 10, 0)
      ) %>%
      group_by(year, site) %>%
      summarise(
        # Temperature summaries
        tmean_season   = mean((T2M_MAX + T2M_MIN) / 2, na.rm = TRUE),
        tmax_mean      = mean(T2M_MAX, na.rm = TRUE),
        tmin_mean      = mean(T2M_MIN, na.rm = TRUE),
        # Heat stress: days where max temp > 35C
        heat_stress_days = sum(T2M_MAX > 35, na.rm = TRUE),
        # Growing degree days (accumulated)
        gdd_total      = sum(gdd, na.rm = TRUE),
        # Precipitation
        precip_total   = sum(PRECTOTCORR, na.rm = TRUE),
        # Solar radiation
        srad_mean      = mean(ALLSKY_SFC_SW_DWN, na.rm = TRUE),
        .groups = "drop"
      )

  }, error = function(e) {
    message("Failed for site=", site, " year=", year, ": ", e$message)
    return(NULL)
  })
}

# Pull weather for all site-years
cat("Pulling weather data... this takes ~5-10 minutes\n")

weather_list <- vector("list", nrow(all_sites))

for (i in seq_len(nrow(all_sites))) {
  cat("Pulling", i, "of", nrow(all_sites),
      "| site:", all_sites$site[i],
      "| year:", all_sites$year[i], "\n")

  weather_list[[i]] <- pull_weather(
    year = all_sites$year[i],
    lat  = all_sites$latitude[i],
    lon  = all_sites$longitude[i],
    site = all_sites$site[i]
  )


  Sys.sleep(0.5)
}

# Combine all results
weather_df <- bind_rows(weather_list)

cat("Weather data pulled successfully!\n")
cat("Rows:", nrow(weather_df), "\n")
cat("Columns:", ncol(weather_df), "\n")

# Save to file
write_csv(weather_df, "../data/weather_season_summary.csv")
cat("Saved to ../data/weather_season_summary.csv\n")
```


## Load weather data and combine with training and testing data
```{r}
#| message: false
#| warning: false

# Load the saved weather file
weather_df = read_csv(here("data", "weather_season_summary.csv"))

# Join weather to training data
train_df = train_df %>%
  left_join(weather_df, by = c("year", "site"))

# Join weather to testing data
test_df = test_df %>%
  left_join(weather_df, by = c("year", "site"))

cat("Training columns after weather join:", ncol(train_df), "\n")
cat("Testing columns after weather join:", ncol(test_df), "\n")

# Check weather NAs
cat("\nWeather NAs in training:\n")
train_df %>%
  select(tmean_season, tmax_mean, tmin_mean, 
         heat_stress_days, gdd_total, precip_total, srad_mean) %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "column", values_to = "n_NA") %>%
  print()

# Update n_pred after adding weather columns
n_pred = train_df %>%
  select(-yield_mg_ha, -site_year) %>%
  ncol()

cat("\nUpdated number of predictors:", n_pred, "\n")
```

# Define a recipe using yield_mg_ha as the outcome
```{r}
#| message: false
#| warning: false

cat("\nCreate recipe for data preprocessing\n")

corn_rec = recipe(
  yield_mg_ha ~ year + site + replicate + block + hybrid + date_planted + 
    date_harvested + grain_moisture + previous_crop + longitude + latitude + soilpH + 
    om_pct + soilk_ppm + soilp_ppm + tmean_season + tmax_mean + tmin_mean + 
    heat_stress_days + gdd_total + precip_total + srad_mean,
  data = train_df
) %>%
  # Convert character predictors to factors
  step_string2factor(all_nominal_predictors()) %>%
  
  # Date variables from the character date columns
  step_mutate(
    planted_date   = mdy(date_planted),
    harvested_date = mdy(date_harvested)
  ) %>%
  
  # Engineer numeric date features
  step_mutate(
    doy_planted      = yday(planted_date),
    doy_harvested    = yday(harvested_date),
    harvest_interval = as.numeric(harvested_date - planted_date)
  ) %>%
  
  # Remove original string and intermediate Date columns
  step_rm(date_planted, date_harvested, planted_date, harvested_date) %>%
  
  # Impute missing numeric and categorical data
  step_impute_median(all_numeric_predictors()) %>%   # numeric median imputation[web:37][web:43]
  step_impute_mode(all_nominal_predictors()) %>%     # mode imputation for factors[web:40]
  
  # Dummy categorical predictors encoded
  step_dummy(all_nominal_predictors())

corn_rec
```

```{r corn_prep}
# Prep the recipe to estimate any required statistics
corn_prep <- corn_rec %>% 
  prep()

# Examine preprocessing steps
corn_prep
```

## **Create grouped resampling**
```{r}
#| message: false
#| warning: false

set.seed(123)

cv_folds = group_vfold_cv(
  train_df,
  group = site_year,
  v = 5
)
cv_folds

```

# Model 1: **XGBoost** 
```{r}
#| message: false
#| warning: false

# XGBoost spec with tunable hyperparameters
xgb_spec = boost_tree(
  # Total number of boosting iterations
  trees          = 1000,
  # Maximum depth of each tree
  tree_depth     = tune(),
  # Step size shrinkage for each boosting step
  learn_rate     = tune(),
  loss_reduction = tune(),
  # Minimum samples required to split a node
  min_n          = tune(),
  sample_size    = tune(),
  mtry           = tune()
) %>%
  set_mode("regression") %>%
  set_engine("xgboost")

xgb_spec
```

# XGBoost workflow
```{r}
#| message: false
#| warning: false

xgb_wf = workflow() %>%
  add_model(xgb_spec) %>%
  add_recipe(corn_rec)

xgb_wf
```

## **Tuning** XGBoost **grid**

```{r}
#| message: false
#| warning: false

set.seed(123)

# Use Latin space‑filling over a range in the tuning
xgb_grid = grid_space_filling(
  tree_depth(),
  learn_rate(range = c(-3, -0.5)),
  trees(),
  loss_reduction(),
  min_n(),
  sample_size = sample_prop(),
  mtry(range = c(5, min(50, n_pred))),  # use the n_pred counted from train_df
  size = 30
)

xgb_grid
```


## Hyperparameter grid visualization
```{r}
# explore how different tuning settings relate to each other

ggplot(data = xgb_grid,
       aes(x = tree_depth, 
           y = min_n)) +
  geom_point(aes(color = factor(learn_rate),
                 size = trees),
             alpha = .5,
             show.legend = FALSE)
```


# Detecting total cores
```{r}
# This help us to detect the total number of available cores and then do a parallel processing on each of them
# Detect Cores for Slurm (Cluster - Sapelo2)

# lets look into our current environment to see the available cores
n_cores <- as.numeric(Sys.getenv("SLURM_CPUS_ON_NODE"))

if (is.na(n_cores)) n_cores <- parallel::detectCores() - 1

# Start the Cluster
cl <- makePSOCKcluster(n_cores)

registerDoParallel(cl)

cat(paste0("\nFound and registered ", n_cores, " cores to work with\n"))

```


# Tune XGBoost with grouped CV folds
```{r}
set.seed(123)

xgb_tuned = tune_grid(
  xgb_wf,
  resamples = cv_folds,
  grid      = xgb_grid,
  metrics   = metric_set(rmse, rsq),
  control   = control_grid(save_pred = TRUE)
)

xgb_tuned

collect_metrics(xgb_tuned) %>%
  filter(.metric == "rmse") %>%
  arrange(mean)
```

# XGB best parameters
```{r}
best_xgb = select_best(xgb_tuned, metric = "rmse")
best_xgb

# Finalize the workflow
final_xgb_wf = finalize_workflow(xgb_wf, best_xgb)
```

# Predicted vs observed on CV predictions
```{r}
xgb_tuned %>%
  collect_predictions() %>%
  inner_join(select_best(xgb_tuned, metric = "rmse")) %>%
  ggplot(aes(x = yield_mg_ha, y = .pred)) +
  geom_point(alpha = 0.3) +
  geom_abline(color = "red") +
  geom_smooth(method = "lm") +
  labs(title = "XGBoost: Predicted vs Observed",
       x = "Observed Yield (mg/ha)",
       y = "Predicted Yield (mg/ha)")
```

# Fit on all training data and predict 2024
```{r}
final_xgb_fit = fit(final_xgb_wf, data = train_df)

xgb_preds = predict(final_xgb_fit, new_data = test_df) %>%
  bind_cols(test_df %>% select(year, site, hybrid)) %>%
  rename(pred_yield_mg_ha = .pred)

xgb_preds
```

## Vip
```{r}
final_xgb_fit %>%
  extract_fit_parsnip() %>%
  vip(num_features = 15) +
  labs(title = "XGBoost Variable Importance")
```


```{r}
# Save final predictions to submission file
test_sub %>%
  mutate(yield_mg_ha = xgb_preds$pred_yield_mg_ha) %>%
  write_csv("../output/testing_submission_filled_xgb.csv")
```

# Model 2: GLMNet
```{r}
#| message: false
#| warning: false
#| 
# GLMNet has only 2 hyperparameters:
# penalty  = how much to shrink coefficients (lambda)
# mixture  = 0 is pure Ridge, 1 is pure Lasso, 0.5 is Elastic Net
glmnet_spec = linear_reg(
  penalty = tune(),
  mixture = tune()
) %>%
  set_engine("glmnet") %>%
  set_mode("regression")

glmnet_spec
```

## GLMNet recipe
```{r}
#| message: false
#| warning: false

# GLMNet REQUIRES normalized predictors (unlike XGBoost)
# So we add step_normalize() here
glmnet_rec = recipe(
  yield_mg_ha ~ year + site + replicate + block + hybrid +
    date_planted + date_harvested +
    grain_moisture +
    previous_crop + longitude + latitude +
    soilpH + om_pct + soilk_ppm + soilp_ppm +
    tmean_season + tmax_mean + tmin_mean +
    heat_stress_days + gdd_total + precip_total + srad_mean,
  data = train_df
) %>%
  # Convert character predictors to factors
  step_string2factor(all_nominal_predictors()) %>%
  
  # Parse date columns
  step_mutate(
    planted_date   = mdy(date_planted),
    harvested_date = mdy(date_harvested)
  ) %>%
  
  # Engineer date features
  step_mutate(
    doy_planted      = yday(planted_date),
    doy_harvested    = yday(harvested_date),
    harvest_interval = as.numeric(harvested_date - planted_date)
  ) %>%
  
  # Remove original date columns
  step_rm(date_planted, date_harvested, planted_date, harvested_date) %>%
  
  # Impute missing values
  step_impute_median(all_numeric_predictors()) %>%
  step_impute_mode(all_nominal_predictors()) %>%
  
  # Dummy encode categorical variables
  step_dummy(all_nominal_predictors()) %>%
  
  # IMPORTANT: GLMNet needs normalized predictors
  step_normalize(all_numeric_predictors())

glmnet_rec
```

## GLMNet workflow
```{r}
#| message: false
#| warning: false

glmnet_wf = workflow() %>%
  add_model(glmnet_spec) %>%
  add_recipe(glmnet_rec)

glmnet_wf
```

## GLMNet tuning grid
```{r}
#| message: false
#| warning: false

set.seed(488)

# Simple grid - only 2 hyperparameters so we can try many combinations
glmnet_grid = grid_regular(
  penalty(range = c(-4, 0)),   # from 0.0001 to 1
  mixture(range = c(0, 1)),    # from pure Ridge to pure Lasso
  levels = 10                  # 10x10 = 100 combinations
)

cat("GLMNet grid size:", nrow(glmnet_grid), "combinations\n")
glmnet_grid
```

## Tune GLMNet with grouped CV
```{r}
set.seed(488)

cv_folds = group_vfold_cv(
  train_df,
  group = site_year,
  v = 5
)
```

```{r}
#| message: false
#| warning: false

# Use same cv_folds as XGBoost for fair comparison
set.seed(488)

# Parallel processing to speed up
library(doParallel)
registerDoParallel(cores = parallel::detectCores() - 1)

glmnet_tuned = tune_grid(
  glmnet_wf,
  resamples = cv_folds,
  grid      = glmnet_grid,
  metrics   = metric_set(rmse, rsq),
  control   = control_grid(save_pred = TRUE)
)

stopImplicitCluster()

cat("GLMNet tuning complete!\n")

# Show best results
collect_metrics(glmnet_tuned) %>%
  filter(.metric == "rmse") %>%
  arrange(mean) %>%
  head(10)
```

## Select best GLMNet model
```{r}
#| message: false
#| warning: false

# Select best model based on lowest RMSE
best_glmnet = glmnet_tuned %>%
  select_best(metric = "rmse")

best_glmnet

# Finalize workflow with best parameters
final_glmnet_wf = finalize_workflow(glmnet_wf, best_glmnet)
final_glmnet_wf
```

## GLMNet predicted vs observed plot
```{r}
#| message: false
#| warning: false

# Collect CV predictions for best model
glmnet_tuned %>%
  collect_predictions() %>%
  inner_join(best_glmnet) %>%
  ggplot(aes(x = yield_mg_ha, y = .pred)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_abline(color = "red", linewidth = 1) +
  geom_smooth(method = "lm", color = "blue") +
  labs(
    title = "GLMNet: Predicted vs Observed Yield",
    x     = "Observed Yield (mg/ha)",
    y     = "Predicted Yield (mg/ha)"
  ) +
  theme_bw()
```

## GLMNet performance metrics
```{r}
#| message: false
#| warning: false

# Show RMSE and R2 for best GLMNet model
glmnet_tuned %>%
  collect_metrics() %>%
  filter(.metric %in% c("rmse", "rsq")) %>%
  inner_join(best_glmnet) %>%
  select(.metric, mean, std_err) %>%
  print()
```

## Fit GLMNet on all training data and predict 2024

```{r}
#| message: false
#| warning: false

# Fit final model on full training data
final_glmnet_fit = fit(final_glmnet_wf, data = train_df)

# Predict 2024 test data
glmnet_preds = predict(final_glmnet_fit, new_data = test_df) %>%
  bind_cols(test_df %>% select(year, site, hybrid)) %>%
  rename(pred_yield_mg_ha = .pred)

glmnet_preds
```

## Save GLMNet predictions to submission file

```{r}
#| message: false
#| warning: false

# Save GLMNet predictions
test_sub %>%
  mutate(yield_mg_ha = glmnet_preds$pred_yield_mg_ha) %>%
  write_csv("../output/testing_submission_glmnet.csv")

cat("GLMNet predictions saved!\n")
```


```{r}
knitr::purl("Final_Project_Group_10.qmd", output = "final_project_group_10_sapelo_script.R", documentation = 0)
```


