---
title: "Final Project"
format: html
editor: visual
---

# Load packages

```{r}
#| message: false
#| message: false

library(tidymodels)
library(tidyverse)
library(recipes)  # step_rm to work
library(lubridate)
library(ranger)
library(rsample)
```

# Loading and inspecting the training and testing data files

```{r}
#| message: false
#| message: false

# list.files("../data/training", full.names = TRUE)
# list.files("../data/testing", full.names = TRUE)

```

## **Read all CSV files**
```{r}
#| message: false
#| warning: false

train_meta  <- read_csv("../data/training/training_meta.csv")
train_soil  <- read_csv("../data/training/training_soil.csv")
train_trait <- read_csv("../data/training/training_trait.csv")

test_meta   <- read_csv("../data/testing/testing_meta.csv")
test_soil   <- read_csv("../data/testing/testing_soil.csv")
test_sub    <- read_csv("../data/testing/testing_submission.csv")
```

## Clean site names and join all files

```{r}
#| message: false
#| warning: false

# The data has messy site names across files that need to match before joining.
# Examples of problems found:
#   trait has: IAH1a, IAH1b, IAH1c  -> should match meta's: IAH1
#   trait has: TXH1-Dry, TXH1-Early -> should match meta's: TXH1
#   meta has:  MOH1- rep 1          -> should match soil's: MOH1
#   soil has:  IAH1_2015            -> should match meta's: IAH1

# This function cleans all of those patterns to a common base site name
clean_site <- function(s) {
  s <- str_remove(s, "_\\d{4}$")          # removes _2015 etc  (soil)
  s <- str_remove(s, "[abc]$")            # removes trailing a/b/c  (trait)
  s <- str_remove(s, "-\\s*(Dry|Early|Late)")   # removes -Dry/-Early/-Late  (trait)
  s <- str_remove(s, "\\s*-?\\s*rep\\s*\\d+")  # removes rep 1 / rep 2  (meta)
  s <- str_remove(s, "_[12]$")            # removes _1 or _2 at end
  str_trim(s)
}

# --- Clean and deduplicate training meta ---
train_meta_clean <- train_meta %>%
  mutate(site = clean_site(site)) %>%
  group_by(year, site) %>%
  summarise(
    previous_crop = first(previous_crop),
    longitude     = mean(longitude, na.rm = TRUE),
    latitude      = mean(latitude,  na.rm = TRUE),
    .groups       = "drop"
  )

# --- Clean and deduplicate training soil ---
train_soil_clean <- train_soil %>%
  mutate(site = clean_site(site)) %>%
  group_by(year, site) %>%
  summarise(
    soilpH    = mean(soilpH,    na.rm = TRUE),
    om_pct    = mean(om_pct,    na.rm = TRUE),
    soilk_ppm = mean(soilk_ppm, na.rm = TRUE),
    soilp_ppm = mean(soilp_ppm, na.rm = TRUE),
    .groups   = "drop"
  )

# --- Clean testing soil ---
test_soil_clean <- test_soil %>%
  mutate(site = clean_site(site)) %>%
  group_by(year, site) %>%
  summarise(
    soilpH    = mean(soilpH,    na.rm = TRUE),
    om_pct    = mean(om_pct,    na.rm = TRUE),
    soilk_ppm = mean(soilk_ppm, na.rm = TRUE),
    soilp_ppm = mean(soilp_ppm, na.rm = TRUE),
    .groups   = "drop"
  )

# --- Clean site in trait before joining ---
train_trait_clean <- train_trait %>%
  mutate(site = clean_site(site))

# --- Build training dataset: trait + meta + soil ---
train_df <- train_trait_clean %>%
  left_join(train_meta_clean,  by = c("year", "site")) %>%
  left_join(train_soil_clean,  by = c("year", "site"))

# --- Build testing dataset: submission + meta + soil ---
test_df <- test_sub %>%
  left_join(test_meta,         by = c("year", "site")) %>%
  left_join(test_soil_clean,   by = c("year", "site"))

# --- Check results ---
cat("Training rows:", nrow(train_df), "\n")
cat("Testing rows:",  nrow(test_df),  "\n")

cat("\nTraining NAs per column:\n")
train_df %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "column", values_to = "n_NA") %>%
  filter(n_NA > 0) %>%
  print()

cat("\nTesting NAs per column:\n")
test_df %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "column", values_to = "n_NA") %>%
  filter(n_NA > 0) %>%
  print()
```

## Add site_year grouping variable

```{r}
#| message: false
#| warning: false

train_df <- train_df %>%
  mutate(site_year = paste(site, year, sep = "_"))

test_df <- test_df %>%
  mutate(site_year = paste(site, year, sep = "_"))
```

## Convert character columns to factors

```{r}
#| message: false
#| warning: false

train_df <- train_df %>%
  mutate(
    site          = factor(site),
    hybrid        = factor(hybrid),
    previous_crop = factor(previous_crop),
    site_year     = factor(site_year)
  )

test_df <- test_df %>%
  mutate(
    site          = factor(site),
    hybrid        = factor(hybrid),
    previous_crop = factor(previous_crop),
    site_year     = factor(site_year)
  )
```

## Count predictors (needed for mtry in XGBoost grid)

```{r}
n_pred <- train_df %>%
  select(-yield_mg_ha, -site_year) %>%
  ncol()

cat("Number of predictors:", n_pred, "\n")
```

# Define a recipe using yield_mg_ha as the outcome

```{r}
#| message: false
#| message: false

corn_rec = recipe(
  yield_mg_ha ~ year + site + replicate + block + hybrid +
    date_planted + date_harvested +
    grain_moisture +
    previous_crop + longitude + latitude +
    soilpH + om_pct + soilk_ppm + soilp_ppm,
  data = train_df
) %>%
  # Convert character predictors to factors
  step_string2factor(all_nominal_predictors()) %>%
  
  # Date variables from the character date columns
  step_mutate(
    planted_date   = mdy(date_planted),
    harvested_date = mdy(date_harvested)
  ) %>%
  
  # Engineer numeric date features
  step_mutate(
    doy_planted      = yday(planted_date),
    doy_harvested    = yday(harvested_date),
    harvest_interval = as.numeric(harvested_date - planted_date)
  ) %>%
  
  # Remove original string and intermediate Date columns
  step_rm(date_planted, date_harvested, planted_date, harvested_date) %>%
  
  # Impute missing numeric and categorical data
  step_impute_median(all_numeric_predictors()) %>%   # numeric median imputation[web:37][web:43]
  step_impute_mode(all_nominal_predictors()) %>%     # mode imputation for factors[web:40]
  
  # Dummy categorical predictors encoded
  step_dummy(all_nominal_predictors())

corn_rec

```

## **Create grouped resampling**

```{r}
#| message: false
#| message: false

set.seed(123)

cv_folds = group_vfold_cv(
  train_df,
  group = site_year,
  v = 5
)

corn_rec
cv_folds

```

## **Tuned XGBoost model with grouped CV**

```{r}
#| message: false
#| message: false

# XGBoost spec with tunable hyperparameters
xgb_spec = boost_tree(
  trees          = 1000,
  tree_depth     = tune(),
  learn_rate     = tune(),
  loss_reduction = tune(),
  min_n          = tune(),
  sample_size    = tune(),
  mtry           = tune()
) %>%
  set_mode("regression") %>%
  set_engine("xgboost")

xgb_spec
```

# XGBoost workflow

```{r}
#| message: false
#| message: false

xgb_wf = workflow() %>%
  add_model(xgb_spec) %>%
  add_recipe(corn_rec)

xgb_wf
```

## **Tuning grid for** XGBoost

```{r}
#| message: false
#| message: false

set.seed(123)

# Use Latin space‑filling over a range in the tuning
xgb_grid = grid_space_filling(
  tree_depth(),
  learn_rate(range = c(-3, -0.5)),
  loss_reduction(),
  min_n(),
  sample_size = sample_prop(),
  mtry(range = c(5, min(50, n_pred))),  # use the n_pred counted from train_df
  size = 30
)

xgb_grid
```

# Tune XGBoost with grouped CV

```{r}
set.seed(123)

xgb_tuned = tune_grid(
  xgb_wf,
  resamples = cv_folds,
  grid      = xgb_grid,
  metrics   = metric_set(rmse, rsq),
  control   = control_grid(save_pred = TRUE)
)

xgb_tuned

collect_metrics(xgb_tuned) %>%
  filter(.metric == "rmse") %>%
  arrange(mean)

```

# XGB best parameters

```{r}
best_xgb = select_best(xgb_tuned, metric = "rmse")
best_xgb

# Finalize the workflow
final_xgb_wf = finalize_workflow(xgb_wf, best_xgb)
```

# Fit on all training data and predict 2024

```{r}
final_xgb_fit = fit(final_xgb_wf, data = train_df)

xgb_preds = predict(final_xgb_fit, new_data = test_df) %>%
  bind_cols(test_df %>% select(year, site, hybrid)) %>%
  rename(pred_yield_mg_ha = .pred)

xgb_preds
```

## 
