---
title: "Final Project"
format: html
editor: visual
---

# Load packages

```{r}
#| message: false
#| message: false

library(tidymodels)
library(tidyverse)
library(recipes)  # step_rm to work
library(lubridate)
library(ranger)
library(rsample)
```

# Loading and inspecting the training and testing data files

```{r}
#| message: false
#| message: false

# list.files("../data/training", full.names = TRUE)
# list.files("../data/testing", full.names = TRUE)

```

## **Read all CSV files**
```{r}
#| message: false
#| warning: false

train_meta  = read_csv("../data/training/training_meta.csv")
train_soil  = read_csv("../data/training/training_soil.csv")
train_trait = read_csv("../data/training/training_trait.csv")

test_meta   = read_csv("../data/testing/testing_meta.csv")
test_soil   = read_csv("../data/testing/testing_soil.csv")
test_sub    = read_csv("../data/testing/testing_submission.csv")
```

## Clean site names and join all files

```{r}
#| message: false
#| warning: false

clean_site = function(s) {
  s <- str_remove(s, "_\\d{4}$")         
  s <- str_remove(s, "[abc]$")            
  s <- str_remove(s, "-\\s*(Dry|Early|Late)")   
  s <- str_remove(s, "\\s*-?\\s*rep\\s*\\d+")  
  s <- str_remove(s, "_[12]$")            
  str_trim(s)
}

# Clean and deduplicate training meta
train_meta_clean = train_meta %>%
  mutate(site = clean_site(site)) %>%
  group_by(year, site) %>%
  summarise(
    previous_crop = first(previous_crop),
    longitude     = mean(longitude, na.rm = TRUE),
    latitude      = mean(latitude,  na.rm = TRUE),
    .groups       = "drop"
  )

# Clean and deduplicate training soil
train_soil_clean = train_soil %>%
  mutate(site = clean_site(site)) %>%
  group_by(year, site) %>%
  summarise(
    soilpH    = mean(soilpH,    na.rm = TRUE),
    om_pct    = mean(om_pct,    na.rm = TRUE),
    soilk_ppm = mean(soilk_ppm, na.rm = TRUE),
    soilp_ppm = mean(soilp_ppm, na.rm = TRUE),
    .groups   = "drop"
  )

# Clean testing soil
test_soil_clean = test_soil %>%
  mutate(site = clean_site(site)) %>%
  group_by(year, site) %>%
  summarise(
    soilpH    = mean(soilpH,    na.rm = TRUE),
    om_pct    = mean(om_pct,    na.rm = TRUE),
    soilk_ppm = mean(soilk_ppm, na.rm = TRUE),
    soilp_ppm = mean(soilp_ppm, na.rm = TRUE),
    .groups   = "drop"
  )

# Clean site in trait before joining
train_trait_clean = train_trait %>%
  mutate(site = clean_site(site))

# Build training dataset: trait + meta + soil
train_df = train_trait_clean %>%
  left_join(train_meta_clean,  by = c("year", "site")) %>%
  left_join(train_soil_clean,  by = c("year", "site"))

# build testing dataset: submission + meta + soil
test_df = test_sub %>%
  left_join(test_meta,         by = c("year", "site")) %>%
  left_join(test_soil_clean,   by = c("year", "site"))

#  Check results 
cat("Training rows:", nrow(train_df), "\n")
cat("Testing rows:",  nrow(test_df),  "\n")

cat("\nTraining NAs per column:\n")
train_df %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "column", values_to = "n_NA") %>%
  filter(n_NA > 0) %>%
  print()

cat("\nTesting NAs per column:\n")
test_df %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "column", values_to = "n_NA") %>%
  filter(n_NA > 0) %>%
  print()
```

## Add site_year grouping variable

```{r}
#| message: false
#| warning: false

train_df = train_df %>%
  mutate(site_year = paste(site, year, sep = "_"))

test_df = test_df %>%
  mutate(site_year = paste(site, year, sep = "_"))
```

## Convert character columns to factors

```{r}
#| message: false
#| warning: false

train_df = train_df %>%
  mutate(
    site          = factor(site),
    hybrid        = factor(hybrid),
    previous_crop = factor(previous_crop),
    site_year     = factor(site_year)
  )

test_df = test_df %>%
  mutate(
    site          = factor(site),
    hybrid        = factor(hybrid),
    previous_crop = factor(previous_crop),
    site_year     = factor(site_year)
  )
```

## Count predictors (needed for mtry in XGBoost grid)

```{r}
n_pred = train_df %>%
  select(-yield_mg_ha, -site_year) %>%
  ncol()

cat("Number of predictors:", n_pred, "\n")
```
## Pull weather data using nasapower

```{r}
#| message: false
#| warning: false
#| eval: false   

library(nasapower)

# Get unique site-year-lat-lon from BOTH training and testing meta
# We need weather for every site-year we have coordinates for
sites_train = train_meta_clean %>%
  filter(!is.na(latitude), !is.na(longitude)) %>%
  select(year, site, latitude, longitude) %>%
  distinct()

sites_test = test_meta %>%
  select(year, site, latitude, longitude) %>%
  filter(!is.na(latitude), !is.na(longitude)) %>%
  distinct()

all_sites = bind_rows(sites_train, sites_test) %>%
  distinct(year, site, .keep_all = TRUE)

cat("Total site-years to pull weather for:", nrow(all_sites), "\n")

# Function to pull growing season weather for one site-year
# Corn growing season = May (day 121) to October (day 304)
pull_weather = function(year, lat, lon, site) {
  
  tryCatch({
    raw = get_power(
      community  = "AG",
      lonlat     = c(lon, lat),
      pars       = c("T2M_MAX",   # daily max temperature (C)
                     "T2M_MIN",   # daily min temperature (C)  
                     "PRECTOTCORR", # precipitation (mm/day)
                     "ALLSKY_SFC_SW_DWN"), # solar radiation (MJ/m2/day)
      dates      = c(paste0(year, "-05-01"),   # May 1
                     paste0(year, "-10-31")),  # Oct 31
      temporal_api = "daily"
    )
    
    raw %>%
      mutate(
        site = site,
        year = year,
        # Growing degree days for corn (base 10C, max 30C)
        tmax_capped = pmin(T2M_MAX, 30),
        tmin_capped = pmax(T2M_MIN, 10),
        gdd = pmax((tmax_capped + tmin_capped) / 2 - 10, 0)
      ) %>%
      group_by(year, site) %>%
      summarise(
        # Temperature summaries
        tmean_season   = mean((T2M_MAX + T2M_MIN) / 2, na.rm = TRUE),
        tmax_mean      = mean(T2M_MAX, na.rm = TRUE),
        tmin_mean      = mean(T2M_MIN, na.rm = TRUE),
        # Heat stress: days where max temp > 35C
        heat_stress_days = sum(T2M_MAX > 35, na.rm = TRUE),
        # Growing degree days (accumulated)
        gdd_total      = sum(gdd, na.rm = TRUE),
        # Precipitation
        precip_total   = sum(PRECTOTCORR, na.rm = TRUE),
        # Solar radiation
        srad_mean      = mean(ALLSKY_SFC_SW_DWN, na.rm = TRUE),
        .groups = "drop"
      )
    
  }, error = function(e) {
    message("Failed for site=", site, " year=", year, ": ", e$message)
    return(NULL)
  })
}

# Pull weather for all site-years
cat("Pulling weather data... this takes ~5-10 minutes\n")

weather_list <- vector("list", nrow(all_sites))

for (i in seq_len(nrow(all_sites))) {
  cat("Pulling", i, "of", nrow(all_sites), 
      "| site:", all_sites$site[i], 
      "| year:", all_sites$year[i], "\n")
  
  weather_list[[i]] <- pull_weather(
    year = all_sites$year[i],
    lat  = all_sites$latitude[i],
    lon  = all_sites$longitude[i],
    site = all_sites$site[i]
  )
  
 
  Sys.sleep(0.5)
}

# Combine all results
weather_df <- bind_rows(weather_list)

cat("Weather data pulled successfully!\n")
cat("Rows:", nrow(weather_df), "\n")
cat("Columns:", ncol(weather_df), "\n")

# Save to file
write_csv(weather_df, "../data/weather_season_summary.csv")
cat("Saved to ../data/weather_season_summary.csv\n")
```

## Load weather and join to training and testing data

```{r}
#| message: false
#| warning: false

# Load the saved weather file
weather_df = read_csv("../data/weather_season_summary.csv")

# Join weather to training data
train_df = train_df %>%
  left_join(weather_df, by = c("year", "site"))

# Join weather to testing data
test_df = test_df %>%
  left_join(weather_df, by = c("year", "site"))

cat("Training columns after weather join:", ncol(train_df), "\n")
cat("Testing columns after weather join:", ncol(test_df), "\n")

# Check weather NAs
cat("\nWeather NAs in training:\n")
train_df %>%
  select(tmean_season, tmax_mean, tmin_mean, 
         heat_stress_days, gdd_total, precip_total, srad_mean) %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "column", values_to = "n_NA") %>%
  print()

# Update n_pred after adding weather columns
n_pred = train_df %>%
  select(-yield_mg_ha, -site_year) %>%
  ncol()

cat("\nUpdated number of predictors:", n_pred, "\n")
```


# Define a recipe using yield_mg_ha as the outcome

```{r}
#| message: false
#| message: false

corn_rec = recipe(
  yield_mg_ha ~ year + site + replicate + block + hybrid +
    date_planted + date_harvested +
    grain_moisture +
    previous_crop + longitude + latitude +
    soilpH + om_pct + soilk_ppm + soilp_ppm +
    tmean_season + tmax_mean + tmin_mean +
    heat_stress_days + gdd_total + precip_total + srad_mean,
  data = train_df
) %>%
  # Convert character predictors to factors
  step_string2factor(all_nominal_predictors()) %>%
  
  # Date variables from the character date columns
  step_mutate(
    planted_date   = mdy(date_planted),
    harvested_date = mdy(date_harvested)
  ) %>%
  
  # Engineer numeric date features
  step_mutate(
    doy_planted      = yday(planted_date),
    doy_harvested    = yday(harvested_date),
    harvest_interval = as.numeric(harvested_date - planted_date)
  ) %>%
  
  # Remove original string and intermediate Date columns
  step_rm(date_planted, date_harvested, planted_date, harvested_date) %>%
  
  # Impute missing numeric and categorical data
  step_impute_median(all_numeric_predictors()) %>%   # numeric median imputation[web:37][web:43]
  step_impute_mode(all_nominal_predictors()) %>%     # mode imputation for factors[web:40]
  
  # Dummy categorical predictors encoded
  step_dummy(all_nominal_predictors())

corn_rec

```

## **Create grouped resampling**

```{r}
#| message: false
#| message: false

set.seed(123)

cv_folds = group_vfold_cv(
  train_df,
  group = site_year,
  v = 5
)

corn_rec
cv_folds

```

## **Tuned XGBoost model with grouped CV**

```{r}
#| message: false
#| message: false

# XGBoost spec with tunable hyperparameters
xgb_spec = boost_tree(
  trees          = 1000,
  tree_depth     = tune(),
  learn_rate     = tune(),
  loss_reduction = tune(),
  min_n          = tune(),
  sample_size    = tune(),
  mtry           = tune()
) %>%
  set_mode("regression") %>%
  set_engine("xgboost")

xgb_spec
```

# XGBoost workflow

```{r}
#| message: false
#| message: false

xgb_wf = workflow() %>%
  add_model(xgb_spec) %>%
  add_recipe(corn_rec)

xgb_wf
```

## **Tuning grid for** XGBoost

```{r}
#| message: false
#| message: false

set.seed(123)

# Use Latin space‑filling over a range in the tuning
xgb_grid = grid_space_filling(
  tree_depth(),
  learn_rate(range = c(-3, -0.5)),
  loss_reduction(),
  min_n(),
  sample_size = sample_prop(),
  mtry(range = c(5, min(50, n_pred))),  # use the n_pred counted from train_df
  size = 30
)

xgb_grid
```

# Tune XGBoost with grouped CV

```{r}
set.seed(123)

xgb_tuned = tune_grid(
  xgb_wf,
  resamples = cv_folds,
  grid      = xgb_grid,
  metrics   = metric_set(rmse, rsq),
  control   = control_grid(save_pred = TRUE)
)

xgb_tuned

collect_metrics(xgb_tuned) %>%
  filter(.metric == "rmse") %>%
  arrange(mean)

```

# XGB best parameters

```{r}
best_xgb = select_best(xgb_tuned, metric = "rmse")
best_xgb

# Finalize the workflow
final_xgb_wf = finalize_workflow(xgb_wf, best_xgb)
```

# Predicted vs observed on CV predictions
```{r}
xgb_tuned %>%
  collect_predictions() %>%
  inner_join(select_best(xgb_tuned, metric = "rmse")) %>%
  ggplot(aes(x = yield_mg_ha, y = .pred)) +
  geom_point(alpha = 0.3) +
  geom_abline(color = "red") +
  geom_smooth(method = "lm") +
  labs(title = "XGBoost: Predicted vs Observed",
       x = "Observed Yield (mg/ha)",
       y = "Predicted Yield (mg/ha)")
```



# Fit on all training data and predict 2024

```{r}
final_xgb_fit = fit(final_xgb_wf, data = train_df)

xgb_preds = predict(final_xgb_fit, new_data = test_df) %>%
  bind_cols(test_df %>% select(year, site, hybrid)) %>%
  rename(pred_yield_mg_ha = .pred)

xgb_preds
```

## Vip
```{r}
library(vip)
final_xgb_fit %>%
  extract_fit_parsnip() %>%
  vip(num_features = 15) +
  labs(title = "XGBoost Variable Importance")
```

```{r}
# Save final predictions to submission file
test_sub %>%
  mutate(yield_mg_ha = xgb_preds$pred_yield_mg_ha) %>%
  write_csv("../output/testing_submission_filled.csv")
```

